---
title: "Demonstration: Regional Skills-Gap Divergence Score for Electricians"
subtitle: "FIT5145 Assignment 3 - Characterising and Analysing Data: Demonstration"
author: "Chinni Raja Ammela (36309591)"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: 
  html_document:
    toc: false
    toc_float: false
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center")
library(readxl)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
```

## 1. Purpose

This notebook demonstrates the feasibility of the regional skills-gap early-warning system proposed in the project report. It uses **Electricians (ANZSCO 3411)** as a worked example occupation, combining:

- **Demand side**: Jobs and Skills Australia's Internet Vacancy Index (IVI), monthly online jobs advertisement counts by occupation and JSA Remoteness classification.
- **Supply side**: NCVER's Total VET students and courses 2025 program completions, for the Electrotechnology (UEE, UTE, UTL) training package, by remoteness region. 

Both datasets are publicly available. Download links and access notes are provided in section 2.

## 2. Data sources

| Dataset | Source | Access | Link |
|---|---|---|---|
| Internet Vacancy Index | Jobs and Skills Australia | Free, no login, direct Excel download | https://www.jobsandskills.gov.au/data/internet-vacancy-index |
| Total VET students and courses 2025: program completions | NCVER DataBuilder | Free, no login, interactive table export | https://www.ncver.edu.au/research-and-statistics/data/databuilder |

**Note on this demonstration's data files:** because NCVER's completions data is exported interactively from DataBuilder (not a static downloadable URL), the exact filtered extract used here (Training package = Electrotechnology, broken down by Remoteness region and Year) has been archived alongside the IVI extract at the project's data repository: **[GitHub raw URL to be inserted once repository is created]*.

## 3. Loading and reshaping the data

### 3.1 Internet Vacancy Index (demand side)
```{r load-ivi}
ivi_path <- "internet_vacancies_anzsco4_occupations_jsa_remoteness_and_northern_australia_classification_-_august_2026.xlsx"

ivi_raw <- read_excel(ivi_path, sheet = "JSA Remoteness", col_names = TRUE)

# Filter to Electricians (ANZSCO 3411), reshape from wide (one column per month)
# to long format, and derive calendar year from the Excel date serial number.
elec_ivi <- ivi_raw %>%
  filter(ANZSCO_CODE == "3411") %>%
  select(-Level, -ANZSCO_CODE, -ANZSCO_TITLE) %>%
  rename(remoteness = `JSA Remoteness`) %>%
  pivot_longer(-remoteness, names_to = "date_serial", values_to = "vacancies") %>%
  mutate(
    date = as.Date(as.numeric(date_serial), origin = "1899-12-30"),
    year = year(date)
  )
glimpse(elec_ivi)
```

### 3.2 Exploratory check: data coverage and seasonality
```{r eda-ivi, fig.width=8, fig.height=3.5}
# Check monthly coverage is complete (no missing months) before aggregating to 
# annual totals - an incomplete year would bias the annual sum downward.
coverage <- elec_ivi %>%
  filter(year >= 2019, year <= 2025) %>%
  count(remoteness, year, name = "n_months")

print(coverage, n = 20)

ggplot(elec_ivi, aes(x = date, y = vacancies, colour = remoteness)) + 
  geom_line() +
  labs(title = "Raw monthly Electrician vacancies, by remoteness", 
       x = NULL, y = "Online job ads", colour = NULL) +
  theme_minimal()
```
Every year from 2019-2025 has a full 12 months of data for both regions, so no interpolation or exclusion is needed before aggregating to annual totals. The raw series shows clear COVID-era volatility (a sharp 2020 dip, then a steep 2021-2022 rebound), which motivates using **year-over-year growth rates** rather than raw levels for comparison with completions - growth rates make the two series comparable despite their very different scales (vacancies are counted in the thousands annually; completions in the tens of thousands).
```{r annual-ivi}
ivi_annual <- elec_ivi %>%
  filter(year >= 2019, year <= 2025) %>%
  group_by(remoteness, year) %>%
  summarise(vacancies = sum(vacancies), .groups = "drop")

ivi_annual
```
2026 is excluded as it is a partial year (data to August only) and would distort a year-over-year growth calculation.

### 3.3 NCVER completions (supply side)

```{r load-ncver}
ncver_path <- "Total_VET_students_and_courses_2025.xlsx"

ncv_raw <- read_excel(ncver_path, sheet = "Pivot Data", col_names = FALSE)
header <- as.character(unlist(ncv_raw[1, ]))
data_rows <- ncv_raw[4:10, ]   # the 7 remoteness breakdown rows
colnames(data_rows) <- header

ncver_long <- data_rows %>%
  rename(training_package = 2, remoteness_ncver = 3) %>%
  select(-1) %>%
  pivot_longer(-c(training_package, remoteness_ncver), names_to = "year", values_to = "completions") %>%
  mutate(year = as.numeric(year), completions = as.numeric(completions))

ncver_long
```

### 3.4 Harmonising geography

IVI classifies regions as a binary **Major City / Regional** split. NCVER's DataBuilder exposes a finer 7-category remoteness classification. To join the two datasets, the NCVER categories are collapsed as follows:

- `Major cities` -> **Major City**
- `Inner regional`, `Outer regional`, `Remote`, `Very remote` -> **Regional**
- `Offshore`, `Not known` -> **excluded** (together ~4.5% of total completions; these do not correspond to any IVI remoteness category and cannot be reliably allocated to either side of the binary split)

```{r harmonise}
ncver_annual <- ncver_long %>%
  mutate(remoteness = case_when(
    remoteness_ncver == "Major cities" ~ "Major City",
    remoteness_ncver %in% c("Inner regional", "Outer regional", "Remote", "Very remote") ~ "Regional",
    TRUE ~ "Exclude"
  )) %>%
  filter(remoteness != "Exclude") %>%
  group_by(remoteness, year) %>%
  summarise(completions = sum(completions), .groups = "drop")

excl_pct <- ncver_long %>%
  mutate(excl = remoteness_ncver %in% c("Offshore", "Not known")) %>%
  summarise(pct = 100 * sum(completions[excl]) / sum(completions))
cat("Share of completions excluded by harmonisation:", round(excl_pct$pct, 1), "%\n")

ncver_annual
```
## 4. Combining demand and supply, and computing the divergence score

```{r combine}
combined <- ivi_annual %>%
  inner_join(ncver_annual, by = c("remoteness", "year")) %>%
  group_by(remoteness) %>%
  arrange(year) %>%
  mutate(
    vacancy_growth    = (vacancies   - lag(vacancies))   / lag(vacancies)   * 100,
    completion_growth = (completions - lag(completions)) / lag(completions) * 100,
    divergence = vacancy_growth - completion_growth
  ) %>%
  ungroup()

combined %>%
  mutate(across(where(is.numeric), ~round(., 1))) %>%
  as.data.frame()
```

The **divergence score** is defined as:

$$\text{divergence}_t = \text{vacancy growth rate}_t - \text{completion growth rate}_t$$

A positive score means vacancy demand grew faster than training completions that year (the gap is widening); a negative score means completions grew faster than vacancies (the gap is narrowing). This is the core novel contribution of the proposed system: existing resources compare demand and supply **levels**, but none compute or track this *rate-of-change* signal over time.

## 5. Visualising the result

```{r plot-indexed, fig.width=9, fig.height=4.5}
indexed <- combined %>%
  group_by(remoteness) %>%
  arrange(year) %>%
  mutate(vac_index = vacancies / first(vacancies) * 100,
         comp_index = completions / first(completions) * 100) %>%
  select(remoteness, year, vac_index, comp_index) %>%
  pivot_longer(c(vac_index, comp_index), names_to = "series", values_to = "index") %>%
  mutate(series = recode(series, vac_index = "Vacancies (IVI)", comp_index = "Completions (NCVER)"))

ggplot(indexed, aes(x = year, y = index, colour = series, linetype = series)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  facet_wrap(~remoteness) +
  labs(title = "Electricians: Vacancy Demand vs. Training Completions (Indexed, 2019 = 100)",
       x = NULL, y = "Index (2019 = 100)", colour = NULL, linetype = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

```{r plot-divergence, fig.width=9, fig.height=4.5}
div_data <- combined %>% filter(!is.na(divergence))

ggplot(div_data, aes(x = year, y = divergence, colour = remoteness)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_line(linewidth = 1) + geom_point(size = 2.5) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dotted", linewidth = 0.8) +
  labs(title = "Divergence Score: Vacancy Growth minus Completion Growth",
       subtitle = "Positive = demand outpacing training supply | Negative = supply catching up",
       x = NULL, y = "Divergence score (percentage points)", colour = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## 6. Trend analysis

```{r trend}
for (r in unique(combined$remoteness)) {
  sub <- combined %>% filter(remoteness == r, !is.na(divergence))
  fit <- lm(divergence ~ year, data = sub)
  s <- summary(fit)
  cat(sprintf("%s: slope = %.2f pp/year, p-value = %.3f, R-squared = %.3f (n = %d)\n",
              r, coef(fit)[2], s$coefficients[2,4], s$r.squared, nrow(sub)))
}
```

## 7. Interpretation

For Electricians, the divergence score follows a consistent pattern in **both** Major City and Regional areas: a large positive spike in 2021 (vacancies rebounded far faster than training completions after the 2020 COVID dip), tapering through 2022-2023, and turning **negative** from 2024 onward (completions now growing faster than vacancies in both regions). The fitted linear trend is negative for both regions (approximately -6.4 to -6.6 percentage points per year), consistent with a mismatch that is **narrowing** rather than widening over the observed period - the opposite direction to a worsening shortage, and a genuinely useful signal that a simple "current gap" comparison would not surface.

**Statistical caveat:** with only six year-over-year observations per region, the trend slopes are not statistically signficant at conventional thresholds (p = 0.52 for Major City, p = 0.31 for Regional), and R-squared is modest. This is expected given the sample size and the real-world volatility of the COVID-affected period. The method is represented here as a **directional, feasibility-level indicator** rather than a statistically validated forecast; a production version of this system would need longer and more granular (e.g., quarterly) series to support stronger statistical claims.

## 8. Feasibility conclusion

This demonstration confirms the core mechanism proposed in the project report is technically feasible with real, publicly available Australian data:

1. Both data sources can be acquired, reshaped, and joined on a common occupation/region/year key.
2. A genuine divergence (rate-of-change) signal - not just a static level comparison - can be computed and visualised.
3. The resulting signal is interpretable and tells a substantively different story than either raw series alone: it correctly identifies that the Electricians training pipeline, after lagging during the COVID-era demand surge, has since caught up and is now outpacing vacancy growth.

A full production system would extend this single-occupation proof of concept across the full ANZSCO occupation set, automate the NCVER DataBuilder extraction, and incorporate quarterly or monthly completions data where available to improve statistical power.